% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

\textit{“It takes a village (to raise a child).”}\\
\rightline{African proverb}

\vspace{5mm}

First and foremost, I would like to acknowledge my co-supervisors John Marioni and Oliver Stegle, for their guidance and support through my entire PhD.
I cannot emphasise enough how much I have learnt in the last four years, and how much I have grown, as a person and as a scientist.
I am thankful to EMBL-EBI and the University of Cambridge for providing an outstanding environment to perform my research - I have been truly spoilt.
I am thankful to my thesis advisory committee, Eileen Furlong and Duncan Odom, and to my collaborators, Ludovic Vallier, Mariya Chhatriwala, Daniel Gaffney, Natsuhiko Kumasaka and Florian Merkle. \\

The help and guidance of several people was instrumental in making my PhD productive and exciting.
A special thank you goes to Marc Jan Bonder and Daniel Seaton, for being the most amazing colleagues and for their continued presence and support, at work and outside, especially in the second part of my PhD. 
Thank you for the morning coffees, the board games, and just for always being there for me. \\

I would also like to thank Paolo Casale, who in the first few months of my PhD introduced and explained to me key concepts within the field of Statistical Genetics; Davis McCarthy for introducing me to and patiently helping me with all things R and single cell RNA-seq related; and Nils Eling for helping me navigate life as an EBI and Cambridge PhD student in those early days.
I would also like to thank Danilo Horta, for his availability to answer all my questions to do with software and maths.
I would like to thank Julie Jerber, Ximena Ibarra-Soria, Arianne Richard, Rebecca Berrens, Na Cai, Rachel Moore and Sara Pulit for teaching me all the biology and genetics I know, and for being fantastic role models for me, whether they are aware or not!
I am especially thankful to Julie with whom I have worked on one of the projects during my PhD, it was a truly wonderful collaboration and I could not have wished for a better partner.
I would also like to thank all other past and present members of the Marioni group, particularly Aaron, Chris, Mike, Rebecca, Shila and Tom. \\

I would like to thank the following people for reading and commenting this thesis: 
Shila Ghazanfar, 
Nils Eling,
Julie Jerber,
% Rebecca Berrens (ch1.2)
Daniel Seaton,
% Marc Jan Bonder (ch2)
Michael Morgan,
Ximena Ibarra-Soria,
% Chris Laumer (ch1,ch5)
% Danila Bredikhin
% Paolo Casale (ch5,4)
% Rachel Moore
Paula Frampton,
Kendall MacMillan,
Oliver Stegle and
John Marioni.

I would also like to thank Nils Eling and Krishna Kumar for this amazing template. \\

% examiners 
% Cecilia Lindgren and Chris Wallace

Finally, I would like to thank Hannah Currant and Elissavet (Elsa) Kentepozidou, for being the best companions in this PhD journey.
From the very first days in Heidelberg, to wedding dress shopping, to discussing work, science, equality, pandemics and everything in between, I could not have asked for two better people to do this with.
I would also like to thank the rest of the amazing community that are the EBI predocs, including the rest of my batch, besides Hannah and Elsa - Ricard, Umberto, Harald and Melike; as well as previous batches especially Nils, Jack, Dani, Hannah, Nadia and Claudia; and following batches, including Rachel, Martina, Conor, Jose. 
Special mention also to Pauline and Joanna, from EMBL Grenoble. \\


On a more personal note, I would like to thank my friends and family, who despite the distance are always there for me, and are my biggest cheerleaders.
I would like to thank Livia, Teresa, Camilla, Sofia and Martina for having been by my side for the last 22+ years.
Giulia for always believing in me and for being with me on the most important day.
Nagiua and Michele, and the rest of the (extended) family, for everything.
I would like to thank my parents, Daniela and Renato, for always encouraging my curiosity and for making me who I am today.
“Grazie mamma, grazie pa!”, cit.
Last but not least, I would like to thank Kendall, for always believing in me and for being the best partner anyone could ask for.

\end{acknowledgements}
