\chapter{List of Publications}

\section{Published papers}

Anna S.E. Cuomo*, Daniel D. Seaton*, Davis J. McCarthy*, Iker Martinez, Marc Jan Bonder, Jose Garcia-Bernardo, Shradha Amatya, Pedro Madrigal, Abigail Isaacson, Florian Buettner, Andrew Knights, Kedar Nath Natarajan, the Hipsci Consortium, Ludovic Vallier, John C. Marioni, Mariya Chhatriwala, Oliver Stegle. 
Single-cell RNA-sequencing of differentiating iPS cells reveals dynamic genetic effects on gene expression. \textit{Nature Communications}, 2020 (* equal contribution).

\section{Accepted manuscripts}

Julie Jerber*, Daniel D. Seaton*, Anna S.E. Cuomo*, Natsuhiko Kumasaka, James Haldane, Juliette Steer, M Patel, D Pearce, M Andersson, Marc Jan Bonder, Ed Mountjoy, Maya Ghoussaini, Madeline A. Lancaster, the HipSci Consortium, John C. Marioni, Florian T. Merkle, Oliver Stegle, Daniel J. Gaffney.
Population-scale single-cell RNA-seq profiling across dopaminergic neuron differentiation. \textit{Nature Genetics}, 2021, in press (* equal contributions). \\

Ricard Argelaguet, Anna S.E. Cuomo, Oliver Stegle, John C. Marioni. 
Computational principles and challenges for the integration of single-cell multi-modal data. \textit{Nature Biotechnology}, 2021, in press. 

\section{Submitted manuscripts}

Anna S.E. Cuomo*, Giordano Alvari*, Christina B. Azodi*, single cell eQTLGen Consortium, Davis J. McCarthy, Marc Jan Bonder.  
Optimising expression quantitative trait locus mapping workflows for single-cell studies. \textit{bioRxiv}, 2021 (* equal contributions). 
% \\

\section{Manuscripts in preparation}

Anna S.E. Cuomo, Danilo Horta, Elissavet Kentepozidou, John C. Marioni, Oliver Stegle.
A linear mixed model framework to identify complex context-specific eQTL using single cell RNA-sequencing.

